---
output: 
  github_document:
    df_print: tibble
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# neon4cast

<!-- badges: start -->
  [![R-CMD-check](https://github.com/eco4cast/neon4cast/workflows/R-CMD-check/badge.svg)](https://github.com/eco4cast/neon4cast/actions)
<!-- badges: end -->

`neon4cast` provides a collection of convenient helper utilities for anyone entering the EFI NEON Forecasting Challenge.  

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("eco4cast/neon4cast")
```

## Examples

```{r}
library(neon4cast)
```

### Validate a forecast file

```{r}
example <- "https://data.ecoforecast.org/forecasts/aquatics/aquatics-2021-02-01-EFInull.csv.gz"
download.file(example, "aquatics-2021-02-01-EFInull.csv.gz")
forecast_output_validator("aquatics-2021-02-01-EFInull.csv.gz")
```

### Compute forecast scores locally

```{r}
forecast <- readr::read_csv("aquatics-2021-02-01-EFInull.csv.gz")
score(forecast, theme = "aquatics")
```

### Generate forecast metadata in EML

Coming soon!

### Access EFI snapshots of NOAA forecasts at NEON sites

```{r}
download_noaa("ABBY")
abby <- stack_noaa()
abby
```

### Submit a forecast

``` r
submit("aquatics-2021-02-01-EFInull.csv.gz")
```

### Other functions

Encountered a bug?  Facing another challenge in participating in the challenge? Developed a cool approach you would like to share with the community? Open an [issue](https://github.com/eco4cast/neon4cast/issues) or [pull request](https://github.com/eco4cast/neon4cast/pulls) here!


```{r include=FALSE}
unlink("aquatics-2021-02-01-EFInull.csv.gz")
```