```{r setup}
library(neon4cast) # remotes::install_github("eco4cast/neon4cast")
library(tidyverse)
library(fable)
```

# Data Access

```{r}
targets <-
  "https://data.ecoforecast.org/targets/beetles/beetles-targets.csv.gz" %>% 
  read_csv(col_types = "cDdd") %>% 
  as_tsibble(index = time, key = siteID)
```

For illustrative purposes, we will pretend most recent year is "future" data. In practice, "past" would be the `targets` data.frame downloaded on the day code is run (i.e. most recent data), while "future" would be the "targets" data downloaded some period of time (i.e. 1 yr) after when it is possible to test if the forecast was accurate.


```{r}
past <-  targets %>% filter(time < max(time) - 365)
future <- targets %>% filter(time >= max(time) - 365)
```

## Compute a forecast

```{r}
## Compute a simple mean/sd model per site... obviously silly given huge seasonal aspect
null_richness <- past  %>% 
  model(null = MEAN(richness)) %>%
  forecast(h = "1 year")

null_abundance <- past  %>%
  model(null = MEAN(abundance)) %>%
  forecast(h = "1 year")
```


## Visualize the forecast

```{r}
first4 <- unique(fc_richness$siteID)[1:4]

null_richness %>% filter(siteID %in% first4)  %>% autoplot(past) + ggtitle("richness")
null_abundance %>% filter(siteID %in% first4)  %>% autoplot(targets) + ggtitle("abundance")
```

## Score the forecast

`fable` can compute CRPS, by default computes averages by siteID (`key`) and `.model` only.  
Optionally, we could include `by = c(".model", "siteID", "time")` to generate separate forecasts by time. 
(For some reason `fable`s internal method is much slower than the EFI code, though results are numerically equivalent). 
Note that fable can work across multiple models, (though we have only one in this case), but not across multiple target variables.
We must handle predictions of `richness` and `abundance` separately.
Recall that smaller CRPS scores are better.
  

```{r}
null_richness %>%
  accuracy(future, list(crps = CRPS))
```

## EFI Formatting

EFI requires a flat-file format for forecasts that avoids the use of complex list columns.  
To convey uncertainty, forecasts must be expressed either by giving mean and standard deviation (for predictions that are normally distributed) or must express forecasts as an ensemble of replicate draws from forecast distribution.
The helper function `efi_format()` handles this transformation.

```{r}
## Combine richness and abundance forecasts. drop the 'model' column
null_forecast <- inner_join(efi_format(null_richness), 
                            efi_format(null_abundance)) %>%
  select(!.model) # we have only one model
```

Score the forecast using EFI's internal method. By default, EFI's method reports the score every unique site-time combination (unique grouping variables).
It is easy to later average across times for a by-site score.

```{r}
scores_null <- neon4cast::score(null_forecast, "beetles")
# average richness scores by site
scores_null %>% group_by(target,siteID) %>% summarise(mean = mean(score, na.rm=TRUE)) %>% filter(target=="richness")
```

# Richer forecast models


A richer forecast may make use of NOAA weather predictions.

```{r}
sites <- unique(targets$siteID)
lapply(sites, neon4cast::download_noaa, dir = neonstore::neon_dir())
noaa <- neon4cast::stack_noaa()
```
In calibrating our model, we can use NEON's local measurements of the meterological variables.

```{r}

# Summary statistics for fixed-location sites:
neonstore::neon_download("DP4.00001.001")
neonstore::neon_store(product = "DP4.00001.001")
# temperature

# humidity

# 


```



```{r}
gapfilled_tagets <- tsibble::fill_gaps(targets)
fc_richness <- gapfilled_tagets  %>% 
  model(arima = ARIMA(richness)) %>%
  forecast(h = "1 year") %>%
  efi_format()
fc_abundance <- gapfilled_tagets  %>%
  model(arima = ARIMA(abundance)) %>%
  forecast(h = "1 year") %>%
  efi_format()

## Combine richness and abundance forecasts. drop the 'model' column
arima_forecast <- inner_join(fc_richness, fc_abundance) %>% select(!.model)
```


Score the forecast

```{r}
arima_scores <- neon4cast::score(arima_forecast, "beetles")
arima_scores %>% group_by(target) %>% summarise(mean = mean(score, na.rm=TRUE))
```
